Night and day: The influence and relative importance of urban characteristics on remotely sensed land surface temperature

### 2020-05-03 report/article.tex
accepted into RSE

% \documentclass[sort&compress]{elsarticle}
% \documentclass[final,3p,times,onecolumn,sort&compress]{elsarticle}
% \documentclass[final,3p,times,twocolumn,sort&compress]{elsarticle}
\documentclass[review]{elsarticle}

@@

%% Packages
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
\usepackage{tabu}
\usepackage{booktabs}
\usepackage{breakurl}
\usepackage{float}
\usepackage{amsmath}
\usepackage{setspace}

@@

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

@@

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}

@@

\title{Night and day: What is the influence and relative importance of urban characteristics on land surface temperature?}

%% or include affiliations in footnotes:
% \author[1]{T.M. Logan\corref{mycorrespondingauthor}}
% \cortext[mycorrespondingauthor]{Corresponding author}
% \ead[url]{www.tomlogan.co.nz}
% \ead{tom.logan@canterbury.ac.nz}

% \author[2]{B. Zaitchik}
% \author[3]{S. Guikema}
% \author[]{A. Nisbet}

% \address[1]{Civil and Natural Resources Engineering, University of Canterbury, New Zealand}
% \address[2]{Earth and Planetary Sciences, Johns Hopkins University, Baltimore, MD}
% \address[3]{Industrial and Operations Engineering, University of Michigan, Ann Arbor, MI}

\begin{abstract}
Heat waves are among the deadliest natural hazards and are expected to increase in frequency and severity under climate change. 
Their impacts in cities can be exacerbated by the urban heat island (UHI), but the mechanisms that underlie severity and timing of the UHI and its interactions with heat waves are poorly understood.
Understanding these mechanisms is necessary to design strategies that reduce land surface temperature and mitigate the effect of heat waves.
Making use of recently available high-resolution day and night thermal satellite imagery and employing advanced nonlinear statistical models, we seek to answer the question \textit{``What is the influence and relative importance of urban characteristics on land surface temperature, during both the day and night?''} 
To answer this question, we analyze urban land surface temperature in four cities across the United States.
In our analysis, we include variables related to vegetation, water, the built-environment, and topography. 
We model the effects of these variables using nonlinear statistical methods which allow for their independent effects to be assessed.
The effects from the daytime and nighttime analysis are compared to determine if previously reported relationships hold between cities and during both the day and night.
Our results suggest that vegetation and impervious surfaces are the most important urban characteristics associated with land surface temperature.
Increasing and decreasing these, respectively, is necessary for reducing high urban temperatures during both night and day.
Our results also demonstrate the potential for using nonlinear statistical analysis to investigate land surface temperature and its relationships with urban characteristics. 
For example, we can evaluate the compounding influence of two urban characteristics on the temperature.
Improved understanding of these relationships influencing both night and day land surface temperature will assist planners undertaking climate change adaptation and heat wave mitigation.
\end{abstract}

\begin{keyword}
LandSat; Land surface temperature; Urban canyon; Machine learning; Convolutional Neural Network
\end{keyword}

@@

\linenumbers

\section{Introduction}
In a warming world, understanding the factors contributing to high land surface temperature (LST) and the urban heat island (UHI) will aid in adapting cities in mitigating urban heat for the health and wellbeing of their communities.
And mitigate they must; the 1995 Chicago heat wave, which killed more than 700 people,\footnote{The 2003 European heatwave killed 70,000 \cite{Robine2008-ky} and the 2015 European heat wave increased mortality up to 30\% \cite{Vicedo-Cabrera2016-si}} is expected to become an annual occurrence by 2080 \cite{klinenberg2015heat}. 
Heat waves' effect on people is exacerbated by the urban heat island (UHI) \cite{Wicki2017-fv, Echevarria_Icaza2016-fr}.
The UHI is understood to be a product of multiple factors including: enhanced absorption of solar radiation, geometric effects that limit radiative cooling and ventilation, anthropogenic heat from vehicles and buildings, air pollution that traps outgoing radiation, high ratios between sensible and latent heat flux due to low vegetation cover and high impervious land cover, and the radiative characteristics of building materials. % \cite{Oke2002-ta, Landsberg1981-mq, Rotach2005-yu}.
The result is increased diurnal and nocturnal temperatures that reduce people's ability to cool off, especially during the night, which drives an increase in mortality \cite{Echevarria_Icaza2016-fr, Murage2017-wj}.

Studies of urban heat include analyses of air temperature, which is a conventional meteorological variable that is often associated with health outcomes(e.g. \cite{Scott2016-lc}), and studies that use satellite-derived land surface temperature (LST) \cite{Imhoff2010-lf, Peng2012-iy, Peng2018-cp, Zhou2014-wc, Voogt2003-mm}.
LST correlates with air temperature at large scales, but there are differences at intra-urban scales \cite{Good2016-yk}.
While LST is not a conventional meteorological variable, it is directly related to the urban heat budget and has the advantage of being available with extensive spatial coverage in gridded satellite products \cite{Hung2006-qy}.
We analyze LST in this study.

Most existing LST studies focus on the daytime \cite{Peng2018-cp,Chun2018-so,Wang2019-water,Zhou2018-iy}.
However, the mechanisms and urban characteristics driving LST allegedly differ between night and day \cite{Hung2006-qy, Chun2017-mm, Nichol2005-mm, Wicki2017-fv, Echevarria_Icaza2016-fr,Sobstyl2018-wt, Peng2012-iy, Zhou2014-wc, Zhao2017-cc}. 
Given that UHI is primarily a nocturnal phenomena, the lack of study on nighttime LST leaves a critical gap in our understanding. 
This can now be addressed as high resolution nighttime satellite images have become available.

Most existing studies that have analyzed LST, particularly nighttime LST, have used low-resolution imagery, making it challenging to deduce a clear understanding of the influence and relative importance of the associated urban characteristics \cite{Chun2017-mm, Echevarria_Icaza2016-fr, Wicki2017-fv, Zhou2014-wc}. 
Additionally, these studies can be enhanced by: 1) studying multiple regions, 2) considering 2D and 3D urban characteristics, 3) diversifying their statistical models and relaxing their linear assumption, 4) rigorously testing and validating their models.
The data availability has limited previous nocturnal studies, meaning that many rely on MODIS images with a 1km resolution \cite{Zhou2014-wc, Echevarria_Icaza2016-fr,Wang2019-tree,Peng2012-iy}.
1km resolution makes it difficult to attribute LST to urban characteristics.
However, 100m resolution LandSat8 (L8) night scenes have recently become available. 
Due to the wide spatial availability of L8 imagery, comparative studies between cities that are necessary to understand the generalizability of findings \cite{Peng2012-iy, Hung2006-qy}, can now be conducted for day and night.

Existing studies have also been criticized for the explanatory variables they've used \cite{Chun2017-mm,Peng2018-cp}.
There is ongoing disagreement regarding the importance of 3D (e.g. building height) vs 2D (such as albedo) variables.  
Competing studies suggest that 3D factors are not important \cite{Berger2017-lx}, while others find that ignoring 3D incorrectly conflates the effect of different 2D variables \cite{Chun2017-mm}.
Beyond the 2D or 3D debate, important variables include green space, water bodies, albedo, and socio-economic factors \cite{Peng2018-cp}. 
However, many studies do not capture these categories and many analyze only the single effect of each variable \cite{Zhao2017-cc, Merbitz2012-xz, Unger2004-ry} (see \cite{Peng2018-cp, Chun2017-mm} for further discussion). 
Considering a variable in isolation, without accounting for potential conflating by other variables, limits the understanding of the interdependent effects that exist.

The third potential enhancement is in the statistical models.
Almost all studies use linear techniques (e.g. \cite{Li2017-yl, Peng2012-iy, Wicki2017-fv,Zhou2014-wc,Peng2018-cp,Echevarria_Icaza2016-fr,Chun2017-mm,Chun2018-so,Wang2019-tree,Wang2019-water}.
Using linear models incurs a number of major challenges that limit these studies' ability to explain the interdependencies between variables. 
The first is that many of the urban characteristics exhibit high multicollinearity \cite{Zhou2014-wc}.
The second is that linear models are limited in their ability to quantify the independent effects of characteristics and their relative influence on LST \cite{Peng2018-cp, Zhou2014-wc}.

Finally, most existing studies do not rigorously validate their models.
The approach they use to assess their accuracy is in-sample validation.
In-sample validation means that model accuracy is assessed with the same data used in the training, rather than unseen data. 
This risks overestimating the accuracy of their models.

The question we address is: \textit{``What is the influence and relative importance of urban characteristics on land surface temperature, during both the day and night?''}
To achieve this we conduct a comparative study of four cities in the United States using nonlinear statistical techniques which capture the interdependencies and relative importance between the urban characteristics.
In addressing our question, we will address the, at times fundamental, limitations of existing studies.
We look at nighttime temperature as well as daytime, we appropriately validate our statistical models, and we use tools that allow us to explore the interdependent, nonlinear, effects of urban characteristics on urban heat.
While we do not claim causality, understanding the associations between urban characteristics on temperatures can complement and further inform our understanding of the processes that lead to high urban land surface temperature.
The results therefore have implications for mitigating the severity of future heat waves.

\section{Data and Methods}
\subsection{Cities studied}

We study four cities in the contiguous United States (Figure \ref{fig:map}): Baltimore, MD; Detroit, MI; Phoenix, AZ; and Portland, OR. These four cities were selected as they include East and West coast cities, a mid-western city, and an arid central city. Phoenix, additionally, has been the subject of numerous other studies on land surface temperature. The cities were also selected due to lidar availability which is required for calculating the 3D variables. The constraint on selecting more cities was the time and computational requirements primarily for the LST and sky view factor calculations.

Figure \ref{fig:map} shows the 500m gridded nighttime LST data for each of the cities and Figure \ref{fig:joy} shows the distribution of each city's 100m resolution day and night land surface temperature.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{fig/report/map_nocturnal_lst.pdf}
    \caption[The nighttime land surface temperature in $^o$C, gridded into 500-meter cells, for the cities studied]{The nighttime land surface temperature in $^o$C, gridded into 500-meter cells, for the cities studied. This is change from the mean (average) for each city.}
    \label{fig:map}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fig/report/joyplot_lst_100.pdf}
    \caption{The distribution of nocturnal and diurnal land surface temperature of the cities studied. 100m resolution.}
    \label{fig:joy}
\end{figure}

\subsection{Land surface temperature}
We calculate land surface temperature using Landsat 8 (Land Processes Distributed Active Archive Center product) imagery, land cover data, and an air temperature observation (data sources provided in \ref{tab:data}). 
Our code is available on our Github\footnote{URL redacted for review} repository, and is as follows:
We convert Band-10 digital-number data to top-of-atmosphere radiance \cite{Jimenez-Munoz2003-wc}. 
We correct for emissivity using land cover data \cite{Alipour2003-ym}, and then calculate the at-satellite brightness temperature \cite{Jimenez-Munoz2003-wc}. 
Finally, an atmospheric correction is made as per the mono-window algorithm \cite{Qin2001-jn} using the maximum observed temperature of the day from a nearby weather station. 
This follows the process is described in \cite{Scott2016-lc}.

To ameliorate the effect of ephemeral changes \cite{Zhou2018-iy} we use at least three minimal-cloud images for each city and night/day period and calculate the mean of LST.

So we can conduct the comparative study between the cities, the land surface temperature for each city is calculated as the difference from the city's mean.

\begin{figure}
    \begin{center}
    \includegraphics[width=\linewidth]{fig/report/scatter_100.png}
    \caption{The land surface temperature in $^o$C of the cities studied at a 100m square resolution.}
    \label{fig:scatter_lst}
    \end{center}
\end{figure}

\subsection{Independent variables}

\textit{Albedo}. Albedo is a measure of the reflectivity or brightness of a surface. It is normalized between 0-100 where the darker surfaces are lower values. Albedo is calculated using the LandSat8 images using the algorithm described in \cite{Smith2010-nw, Liang2001-jd}.

@@

% \textit{Building height}. Building height is estimated using the building footprints and available lidar data following \cite{Chun2017-mm}. The mean lidar elevation within each building's footprint is subtracted by the mean digital elevation (topography) within the footprint to estimate the building's height.

\textit{Elevation}. 1/3 arc-second (~10m) bare-earth elevation (topography) data is available courtesy the U.S. Geological Survey.

\textit{Surface elevation}. The surface elevation is determined from the lidar data. Surface elevation captures the natural and built features.

\textit{Impervious surface percentage}. This is provided in the National Land Cover Database from the Multi-Resolution Land Charasteristic Consortium \cite{Xian2011-aa}. 
To generate the impervious surface area (ISA), the consortium uses LandSat data, NLCD land cover, and nighttime light imagery. 
The stable nighttime light intensity is only used to estimate the boundary of urban areas. 
The Landsat images were converted to top-of-atmosphere reflectance. 
The data is provided at a 30m resolution.

\textit{Land cover}. 
Classification of land cover is described in \cite{Homer2015-ce}. 
However, as it is used in the calculation of emissivity when calculating the LST it cannot be used as an independent variable. 
The only landcover that we use is category 11, water.

\textit{NDBI}. 
Normalized difference built-up index indicates the intensity of imperviousness \cite{Bhatti2014-ae}. 
It is calculated from satellite images as $$NDBI=\frac{B_{SWIR}-B_{NIR}}{B_{SWIR}+B_{NIR}}$$ where $B_{NIR}$ and $B_{SWIR}$ are the reflectances in the near-infrared and short-wave infrared bands respectively \cite{Alhawiti2016-wv}. 
Using LandSat8 imagery, this is $$NDBI=\frac{B6-B5}{B6+B5}$$ \cite{barsi2014}.
% *Kaplan 2018, Peng 2018

\textit{NDVI}. 
The normalized difference vegetation index (NDVI) measures green vegetation. 
It is calculated from satellite images as $$NDVI=\frac{B_{NIR}-B_{red}}{B_{NIR}+B_{red}}$$ where $B_{NIR}$ and $B_{red}$ are the reflectances in the near-infrared and red bands respectively \cite{Alhawiti2016-wv}. 
Using LandSat8 imagery, this is $$NDVI=\frac{B5-B4}{B5+B4}$$ \cite{barsi2014}.

% \textit{Nighttime light intensity}. Stable nighttime light intensity is available from the Defense Meteorological Satellite Program (DMSP) Operational Linescan System (OLS). They prepare the data to remove clouds and ephemeral light sources. We use 2013 (the most recently available) data of version 4 at a spatial resolution of 30 arc-seconds (1km at the equator) \cite{ngdc2013version}. The nighttime light intensity was found to be only in moderate agreement with density \cite{bagan2018}, meaning lights may indicate anthropogenic energy use during the night. However, it's lower resolution means that can not capture minor differences. We resampled the raster to a 30m resolution, as per the NLDC \cite{Homer2015-ce}, and for each grid calculated the mean, min, and max.

\textit{Population density}. 
Some studies have found that population density has a positive influence on LST \cite{Li2017-yl, Peng2018-cp}, although this may be the result of confounding with other factors. 
We calculate population density using the U.S. census at the block level. 
The most recent census was 2010, so we use that data as an estimator of where people reside in the evening. 
This data is also at a lower resolution that the grid cells used, so the grid cell assumes the density of the block that its centroid is contained within.

\textit{Sky view factor}. Urban canyons have been found to have an effect on UHI because they prevent air circulation and hinder nighttime radiative cooling \cite{Landsberg1981-mq, Chun2017-mm} and are used to indicate radiation flux within complex environments \cite{Matzarakis2007-xy}. We calculate SVF using the R packages \textit{horizon} \cite{Van_doninck2018-ib} that is based off the algorithm presented in \cite{Dozier1990-kn}. 
We use the same parameters used by \cite{Chun2017-mm}: the number of search directions, $\phi=10^o$; and the radius of the reference circle, $R=300m$. 
We use a spatial resolution of 6 meters.
% * Yuan 2011, Unger 2004.

\textit{Tree canopy cover}. 
The 2011 edition of percent tree canopy cover is calculated using National Agriculture Imagery Program (NAIP) aerial imagery, Landsat 5 imagery, elevation, and existing NLCD data \cite{Coulston2012-uu, Homer2015-ce}. 
The data provided is at a 30m resolution. 
The six reflective bands from Landsat 5 are used to calculate top-of-atmosphere reflectance \cite{Coulston2012-uu} so the data does not contain information used in the LST calculation (which requires radiance). 
% * Rogan 2013, Elmes 2017

\subsection{Data preparation and robustness}

The objective of our study is to determine the most influential urban characteristics and understand how they relate to land surface temperature.
To do this, we use partial dependence plots to understand how the urban characteristics are associated with land surface temperature.
For a selected characteristic, partial dependence is calculated by fixing the value of that characteristic in all observations in the dataset and predicting the land surface temperature. 
This is repeated over a range for the feature of interest.
To evaluate how multiple urban characteristics interact, partial dependence can also be conducted in multiple dimensions.
The swing \cite{Shortridge2015-ub} measures the relative importance of variables by calculating the magnitude of change in temperature associated with each variable.
To calculate partial dependence, we need a data set and a statistical model.

>>>>

% \documentclass[sort&compress]{elsarticle}
\documentclass[final,3p,times,onecolumn,sort&compress]{elsarticle}
% \documentclass[final,3p,times,twocolumn,sort&compress]{elsarticle}
% \documentclass[review]{elsarticle}

@@

%% Packages
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
\usepackage{tabu}
\usepackage{booktabs}
\usepackage{breakurl}
\usepackage{float}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{array}
\newenvironment{conditions}
  {\par\vspace{\abovedisplayskip}\noindent\begin{tabular}{>{$}l<{$} @{${}={}$} l}}
  {\end{tabular}\par\vspace{\belowdisplayskip}}
\usepackage{listings}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  frame=single,
  breaklines=true,
  postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space},
}

@@

%% APA style
\bibliographystyle{model5-names}\biboptions{authoryear}

@@

%% `Elsevier LaTeX' style
% \bibliographystyle{elsarticle-num}

@@

\title{Night and day: The influence and relative importance of urban characteristics on remotely sensed land surface temperature}

%% or include affiliations in footnotes:
\author[1]{T.M. Logan\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead[url]{www.tomlogan.co.nz}
\ead{tom.logan@canterbury.ac.nz}

\author[2]{B. Zaitchik}
\author[3]{S. Guikema}
\author[4]{A. Nisbet}

\address[1]{Civil and Natural Resources Engineering, University of Canterbury, New Zealand}
\address[2]{Earth and Planetary Sciences, Johns Hopkins University, Baltimore, MD}
\address[3]{Industrial and Operations Engineering, University of Michigan, Ann Arbor, MI}
\address[4]{www.ajnisbet.com}

\begin{abstract}
The characteristics of urban land surfaces contribute to the urban heat island, and, in turn, can exacerbate the severity of heat wave impacts.
However, the mechanisms and complex interactions in urban areas underlying land surface temperature are still being understood.
Understanding these mechanisms is necessary to design strategies that mitigate land temperatures in our cities.
Using the recently available night-time moderate-resolution thermal satellite imagery and employing advanced nonlinear statistical models, we seek to answer the question \textit{``What is the influence and relative importance of urban characteristics on land surface temperature, during both the day and night?''} 
To answer this question, we analyze urban land surface temperature in four cities across the United States.
We devise techniques for training and validating nonlinear statistical models on geostatistical data and use these models to assess the interdependent effects of urban characteristics on urban surface temperature.
Our results suggest that vegetation and impervious surfaces are the most important urban characteristics associated with land surface temperature.
While this may be expected, this is the first study to quantify this relationship for Landsat-resolution nighttime temperature estimates.
Our results also demonstrate the potential for using nonlinear statistical analysis to investigate land surface temperature and its relationships with urban characteristics. 
Improved understanding of these relationships influencing both night and day land surface temperature will assist planners undertaking climate change adaptation and heat wave mitigation.
\end{abstract}

\begin{keyword}
LandSat; Land surface temperature; Urban canyon; Geospatial machine learning; Convolutional neural network; Spatial random forest
\end{keyword}

@@

% \linenumbers

\section{Introduction}
In a warming world, understanding the factors contributing to high land surface temperature (LST) will aid in adapting cities in mitigating urban heat for the health and wellbeing of their communities.
And mitigate they must; the 1995 Chicago heat wave, which killed more than 700 people,\footnote{The 2003 European heatwave killed 70,000 \citep{Robine2008-ky} and the 2015 European heat wave increased mortality up to 30\% \citep{Vicedo-Cabrera2016-si}} is expected to become an annual occurrence by 2080 \citep{klinenberg2015heat}. 
Heat waves' effect on people is exacerbated by the urban heat island (UHI) \citep{Wicki2017-fv, Echevarria_Icaza2016-fr}.
Both the air and surface UHIs are understood to be a product of multiple factors including: enhanced absorption of solar radiation, geometric effects that limit radiative cooling and ventilation, anthropogenic heat from vehicles and buildings, air pollution that traps outgoing radiation, high ratios between sensible and latent heat flux due to low vegetation cover and high impervious land cover, and the radiative characteristics of building materials. % \citep{Oke2002-ta, Landsberg1981-mq, Rotach2005-yu}.
The result is increased diurnal and nocturnal temperatures that reduce people's ability to cool off, especially during the night, which drives an increase in mortality \citep{Echevarria_Icaza2016-fr, Murage2017-wj}.

Studies of urban heat include analyses of air temperature (e.g. \cite{Scott2016-lc}) and satellite-derived land surface temperature (LST) \citep{Imhoff2010-lf, Peng2012-iy, Peng2018-cp, Zhou2014-wc, Voogt2003-mm}.
LST correlates with air temperature at large scales, but there are differences at intra-urban scales \citep{Good2016-yk}.
While LST is not a conventional meteorological variable, it is directly related to the urban heat budget and has the advantage of being available with extensive spatial coverage in gridded satellite products \citep{Hung2006-qy}.
Most existing LST studies focus on the daytime \citep{Peng2018-cp,Chun2018-so,Wang2019-water,Zhou2018-iy} and use the LST to infer the surface UHI (SUHI) \cite{Zhou2018-iy}.
However, the mechanisms and urban characteristics driving LST allegedly differ between night and day \citep{Hung2006-qy, Chun2017-mm, Nichol2005-mm, Wicki2017-fv, Echevarria_Icaza2016-fr,Sobstyl2018-wt, Peng2012-iy, Zhou2014-wc, Zhao2017-cc}. 
Given that night temperatures have a major effect on health (as high night temperatures prevent the human body from recovering from daytime heat exposure \cite{Laaidi2012-ir}), the lack of study on nighttime LST leaves a critical gap in our understanding. 
Ideally, to study this phenomenon satellite imagery that has both higher spatial resolution and captures an area at regular intervals over a 24-hour period would be available; this would enable analysis to understand the heating and cooling processes that lead to the UHI.
While the time-series images are currently not available at scale, higher resolution nighttime satellite images have recently become available and these enable us to study the nighttime LST.

There are an increasing number of studies that use satellite remote sensing to analyze land surface temperature (see \cite{Zhou2018-iy} for a review).
We enhance these studies in the following ways:
\begin{enumerate}
    \item We use surface temperature data at night and day at 100-meter resolution.
    Data availability has limited previous nocturnal studies, meaning that many rely on MODIS images with a 1km resolution \citep{Zhou2014-wc, Echevarria_Icaza2016-fr,Wang2019-tree,Peng2012-iy}.
    1km resolution makes it difficult to attribute LST to urban characteristics \citep{Chun2017-mm, Echevarria_Icaza2016-fr, Wicki2017-fv, Zhou2014-wc}. 
    However, 100m resolution LandSat8 (L8) night scenes have recently become available.

    \item We study four cities in the USA.
    Due to the wide spatial availability of L8 imagery, comparative studies between cities that are necessary to understand the generalizability of findings \citep{Peng2012-iy, Hung2006-qy}, can now be conducted for day and night.

    \item We include both 2D and 3D urban characteristics in our analysis.
    Existing studies have been criticized for the explanatory variables they've used \citep{Chun2017-mm,Peng2018-cp}.
    One point of disagreement is regarding the importance of 3D (e.g. building height) vs 2D (such as albedo) variables.
    Competing studies suggest that 3D factors are not important \citep{Berger2017-lx}, while others find that ignoring 3D incorrectly conflates the effect of different 2D variables \citep{Chun2017-mm}.
    Beyond the 2D or 3D debate, important variables include green space, water bodies, albedo, and socio-economic factors \citep{Peng2018-cp}. 
    However, many studies do not capture all of these categories and many analyze only the univariate effect of each variable: e.g., from \cite{Peng2018-cp} (p.256) ``many studies used only a single influencing factor to establish the regression model with LST, and compare the individual effects of different factors on LST based on the coefficient of regression.'' (see \cite{Peng2018-cp, Chun2017-mm} for further discussion). 
    Considering a variable in isolation, without accounting for potential conflating by other variables, limits the understanding of the interdependent effects that exist.

    \item We use a variety of statistical models that have the ability to capture nonlinear effects and the compounding effects of variables.
    Almost all studies use linear techniques (e.g. \cite{Li2017-yl, Peng2012-iy, Wicki2017-fv,Zhou2014-wc,Peng2018-cp,Echevarria_Icaza2016-fr,Chun2017-mm,Chun2018-so,Wang2019-tree,Wang2019-water, Galletti2019-wa}.
    Using linear models can be much simpler than the alternatives, but it incurs limitations on the models' ability to explain the interdependencies between variables. 
    One limitation is that many of the urban characteristics exhibit high multicollinearity \citep{Zhou2014-wc}.
    A second issue is that linear models are limited in their ability to quantify the independent effects of characteristics and their relative influence on LST \citep{Peng2018-cp, Zhou2014-wc}.
    In achieving this we devise a method for training a random forest on spatially auto-correlated data and use a convolution neural network in a geospatial context for the first time to our knowledge.
    We also use these different statistical models to present the model uncertainty in the effect of each factor.

    \item We rigorously validate our statistical models.
    Many existing studies do not rigorously validate their models.
    The most common approach used to assess statistical accuracy is in-sample validation.
    In-sample validation means that model accuracy is assessed with the same data used in the training, rather than unseen data \cite{Shmueli2011-wd}.
    This risks overestimating the accuracy of their models.
    Instead, we test our models on data unused during the model training and subset this data to avoid overfitting.
\end{enumerate}

With these enhancements, we seek to answer the question: \textit{``What is the influence and relative importance of urban characteristics on land surface temperature, during both the day and night?''}
To achieve this we conduct a comparative study of four cities in the United States using nonlinear statistical techniques which capture the interdependencies and relative importance between the urban characteristics.
We look at nighttime temperature as well as daytime, we appropriately validate our statistical models, and we use tools that allow us to explore the interdependent, nonlinear, effects of urban characteristics on urban heat.
While we do not claim causality, understanding the associations between urban characteristics on temperatures can complement and further inform our understanding of the processes that lead to high urban land surface temperature.
The results therefore have implications for mitigating the severity of future heat waves and preparing our urban regions for a warmer climate.

\section{Data and Methods}
\subsection{Cities studied}
We study four cities in the contiguous United States (Figure \ref{fig:map}): Baltimore, MD; Detroit, MI; Phoenix, AZ; and Portland, OR. 
These four cities were selected as they include East and West coast cities, a mid-western city, and an arid central city. 
Phoenix, additionally, has been the subject of numerous other studies on land surface temperature. 
The cities were also selected due to lidar availability which is required for calculating the 3D variables. 
The constraint on selecting more cities was the time and computational requirements primarily for the LST and sky view factor calculations.

\subsection{Land surface temperature (dependent variable)}
We calculate land surface temperature using Landsat 8 (Land Processes Distributed Active Archive Center product) imagery, land cover data, and an air temperature observation (data sources provided in \ref{tab:data}). 
Our code is available on our Github\footnote{URL redacted for review} repository, and follows the process described in \cite{Scott2016-lc}:
\begin{itemize}
    \item 
        Convert Band-10 digital-number data to top-of-atmosphere radiance \citep{Jimenez-Munoz2003-wc}: 
        \begin{equation}
        \label{eqn:convert_TOA}
        L_\lambda = M_L Q_{cal} + A_L
        \end{equation}
        where:
        \begin{conditions}
         L_\lambda & Top-of-atmosphere spectral radiance \\
         M_L &  Band-specific multiplicative rescaling factor from the satellite metadata \\   
         A_L &  Band-specific additive rescaling factor from the satellite metadata \\   
         Q_{cal} &  Quantized and calibrated standard product pixel values
        \end{conditions}
    \item Determine the emissivity using land cover data \citep{Alipour2003-ym}, and then calculate the at-satellite brightness temperature \citep{Jimenez-Munoz2003-wc}: 
        \begin{equation}
        \label{eqn:at_sat}
        T_s = \frac{K_2}{ln(1 + \frac{K_1}{L})}
        \end{equation}
        where:
        \begin{conditions}
         T_s & At-satellite brightness temperature \\
         L &  Spectral radiance ($L_\lambda$) divided by the emissivity \\   
         K_1, K_2 & Band-specific thermal conversion constants from the satellite metadata
        \end{conditions}
    \item Correct the brightness temperature for atmospheric water vapor using the mono-window algorithm \citep{Qin2001-jn} using the maximum observed temperature of the day from a nearby weather station (see the code in \ref{appendix:code}).
\end{itemize}
The result is a geospatial dataset of land surface temperature at a 100-meter resolution.

To ameliorate the effect of ephemeral changes \citep{Zhou2018-iy} we use at least three minimal-cloud images for each city and night/day period and calculate the mean of LST.
The time and dates for the images used are shown in Table \ref{tab:satellite_meta}.

So we can conduct the comparative study between the cities, the land surface temperature is normalized by subtracting the mean (average) value of that city, i.e., we analyze the `difference from the mean' as the response variable for this study.

\subsection{Independent variables}
The independent variables (i.e., the urban characteristics included to analyze their relationship with land surface temperature) were selected to capture the important factors including green space, water bodies, albedo, and socio-economic factors \citep{Peng2018-cp}. 
The data sources are listed in Table \ref{tab:data}.
Note that these variables are remotely sensed and therefore our study relies on these remote sensing techniques.
Because these techniques are established and from well-regarded data sources, we do not include additional ground truth assessments for these parameters.

\textit{Albedo}. 
Albedo is a measure of the reflectivity or brightness of a surface. 
It is normalized between 0-100 where the darker surfaces have lower values. 
Albedo is calculated using the LandSat8 images using the algorithm described in \cite{Smith2010-nw, Liang2001-jd}. 
These data are at a 100-meter resolution and are calculated using the daytime satellite images.

@@

\textit{Elevation}. 
1/3 arc-second (~10m) bare-earth elevation (topography) data are available courtesy the U.S. Geological Survey.

\textit{Surface elevation}. 
The surface elevation is determined from the lidar data. 
Surface elevation captures the natural and built features. 
The lidar data was aggregated to calculate the surface elevation at a 6-meter resolution.

\textit{Impervious surface percentage}. 
This is provided in the National Land Cover Database from the Multi-Resolution Land Characteristic Consortium \citep{Xian2011-aa}. 
To generate the impervious surface area (ISA), the consortium uses LandSat data, NLCD land cover, and nighttime light imagery. 
The stable nighttime light intensity is only used to estimate the boundary of urban areas. 
The Landsat images were converted to top-of-atmosphere reflectance. 
These data are provided at a 30m resolution.

\textit{Land cover}. 
Classification of land cover is described in \citep{Homer2015-ce}. 
However, as land cover is used when calculating the LST it cannot be used as an independent variable.
That is, a variable used to construct a dependent variable should not be used as a predictor in a statistical model; 
if it were, the statistical model will likely find the mathematical relationship used to create the response, rather than finding the biophysical association we are looking for. 
Additionally, using the same variable in both construction and prediction will likely artificially inflate the stated accuracy of the statistical model due to overfitting.
However, due to the significant biophysical difference between water and other land classifications, and given that the potential to overfit when only one of the categories is included is low, we do include water as an independent variable for the statistical analysis.
These data are provided at a 30m resolution.

\textit{NDBI}. 
Normalized difference built-up index indicates the intensity of imperviousness \citep{Bhatti2014-ae}. 
It is calculated from satellite images as 
\begin{equation}
    \label{eqn:ndbi}
    NDBI = \frac{B_{SWIR}-B_{NIR}}{B_{SWIR}+B_{NIR}} = \frac{B6-B5}{B6+B5}
\end{equation}
where $B_{NIR}=B_5$ and $B_{SWIR}=B_6$ are the reflectances in the near-infrared and short-wave infrared bands respectively \citep{Alhawiti2016-wv, barsi2014}.
These data are at a 100-meter resolution and are calculated using the daytime satellite images.

\textit{NDVI}. 
The normalized difference vegetation index (NDVI) measures green vegetation. 
It is calculated from satellite images as 
\begin{equation}
    \label{eqn:ndvi}
    NDVI = \frac{B_{NIR}-B_{red}}{B_{NIR}+B_{red}} = \frac{B5-B4}{B5+B4}
\end{equation}
where $B_{NIR} = B_5$ and $B_{red} = B_4$ are the reflectances in the near-infrared and red bands respectively \citep{Alhawiti2016-wv, barsi2014}.
These data are at a 100-meter resolution and are calculated using the daytime satellite images.

\textit{Population density}. 
Some studies have found that population density has a positive influence on LST \citep{Li2017-yl, Peng2018-cp}, although this may be the result of confounding with other factors. 
We calculate population density using the U.S. census at the block level. 
The most recent census was 2010, so we use that data as an estimator of where people reside in the evening. 
These data are also at a lower resolution than the grid cells used, so the grid cell assumes the density of the block that its centroid is contained within.
These data are at the neighborhood block level.

\textit{Sky view factor}. 
Urban canyons, which consist of the air volume and surfaces bounded by walls of buildings separated by a street \citep{Oke2017-or}, control the airflow and energy exchange between the ground and roof \citep{Oke1976-zn} so affect the air temperature within the `urban canopy layer.'
This can intensify surface temperatures \citep{Landsberg1981-mq, Chun2017-mm, Matzarakis2007-xy}. 
We calculate SVF using the R package \textit{horizon} \citep{Van_doninck2018-ib} that is based off the algorithm presented in \citep{Dozier1990-kn}. 
We use the same parameters used by \citep{Chun2017-mm}: the number of search directions, $\phi=10^o$; and the radius of the reference circle, $R=300m$. 
We use a spatial resolution of 6 meters.
% * Yuan 2011, Unger 2004.

\textit{Tree canopy cover}. 
The 2011 edition of percent tree canopy cover is calculated using National Agriculture Imagery Program (NAIP) aerial imagery, Landsat 5 imagery, elevation, and existing NLCD data \citep{Coulston2012-uu, Homer2015-ce}. 
The data provided are at a 30m resolution. 
The six reflective bands from Landsat 5 are used to calculate top-of-atmosphere reflectance \citep{Coulston2012-uu} so the data do not contain information used in the LST calculation (which requires radiance). 
% * Rogan 2013, Elmes 2017
These data are available at the 30m resolution.

\subsection{Data processing and variable selection}
To analyze the data from these disparate sources, it needs to be transformed into a suitable format that is consistent between the variables.
The raw data are a variety of spatial data types from ``area level" (the population density data) and ``geostatistical raster" (e.g. the land surface temperature).
We chose to resample the data into a grid of square cells.
We conduct the resampling twice to produce two datasets. The first has 100-meter square cells, the other has 500-meter cells. 
Having two resolutions allows us to evaluate whether our conclusions are sensitive to the spatial resolution used.
For each of the cells the mean, maximum, minimum, and standard deviation of all variables were calculated. 
To address the spatial effects of each variable, the mean of the surrounding cells (including diagonal) was calculated and the resulting `spatial lag' variable was included as an additional independent variable.
